Multi-redshift limits on the 21cm power spectrum from PAPER

### 2014-08-04 ms.tex
compiles for aastex
revised Trms section. now with more clarity

%\documentclass[fleqn,10pt,iop]{emulateapj}
\documentclass{aastex}

@@

@@

\title{Multi-redshift limits on the 21cm power spectrum from PAPER}
\author{
Daniel C. Jacobs\altaffilmark{1},
Aaron R. Parsons\altaffilmark{2,8},
James E. Aguirre\altaffilmark{3},
Zaki Ali\altaffilmark{2},
Judd Bowman\altaffilmark{1},
Richard F. Bradley\altaffilmark{4,5,6},
Chris L.  Carilli\altaffilmark{7,10},
David R. DeBoer\altaffilmark{8},
Matthew R. Dexter\altaffilmark{8},
Nicole E. Gugliucci\altaffilmark{5},
Pat Klima\altaffilmark{5},
Adrian Liu\altaffilmark{2,11},
Dave H. E. MacMahon\altaffilmark{8}
Jason R. Manley\altaffilmark{9},
David F. Moore\altaffilmark{3},
Jonathan C. Pober\altaffilmark{12},
Irina I. Stefan\altaffilmark{10},
William P. Walbrugh\altaffilmark{9}}

@@

\begin{abstract}
We present new observations from the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER) telescope which place new deeper limits on the HI power spectrum over the redshift range of $7.5<z<10.5$, extending previously published single redshift results to cover the full range accessible to the instrument. The epoch of reionization power spectrum is expected to evolve strongly with redshift, and it is this variation with cosmic history that will allow us to begin to place constraints on the physics of reionization.  The primary obstacle to this goal are bright foregrounds and the attendant systematics associated with high dynamic range measurements. Here we present an integration across a full season of data and find systematics just beginning to dominate, though well below any levels previously reported. This  500 hour integration is the longest such yet recorded and tests foreground removal to a theoretical dynamic range of 10,000 to 1. We use filtering techniques that take advantage of the large instrumental bandwidth to make a high dynamic range subtraction of foreground power by averaging together many low sensitivity measurements.  .
 Power spectra at different points across the redshift range reveal the variable efficacy of the foreground subtraction.  Noise limited measurements of $\Delta^2$ at $k=$0.2hMpc$^{-1}$ and z$=7.55$ reach as low as (48mK)$^2$ ($1\sigma$).  Most spectra demonstrate a robust foreground removal of foreground signals to the thermal noise limit, leaving systematics on the scale of the noise.  To better understand the source of these residual systematics we examine different estimates of the noise in the data and compare with theoretical predictions. We find that in the case of the highest redshift power spectrum, foregrounds begin to dominate over the noise, reflecting the increased difficulty in filtering foregrounds near the edge of the instrumental band. 
\end{abstract}

@@

 \begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/psa32_flagging_zbins.png}
\caption{\label{fig:bins} The average amount of data remaining after interference flagging over the 3 month period between Dec 2011 and March 2012 (black line) is quite high.  Redshift bins (in grey, redshift center indicated with text label) are chosen to include spectral channels with uniform weight, i.e. no missing channels while maximizing coverage over the band.  Redshift 7.68 is included for comparison with P14.  Channels with no missing data are indicated by the dotted line, visible at the edge of flagged channels. Each redshift bin is 20 MHz wide, but weighted by a Blackman-Harris window function which heavily down-weights the outer 10MHz for a Noise Equivalent Bandwidth of 10MHz.  The interference is almost exclusively dominated by two features: ORBCOMM satellites at 137MHz and an unidentified intermittent line emitter at 175MHz. The roll off at 115MHz is due the rising noise at low frequencies being incorrectly flagged as interference.  }
\end{figure*} 
%The success of the remainder of the processing stages hinges on the spectral stability of the data, which can easily be assessed by comparing redundant baselines. In figure XXX we see the r.m.s. of the ratios between redundant calibrations, before and after the calibration has been applied.  
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/psa32_trms_with_and_without_fr_filtering_z.png}
\caption{\label{fig:noise} Root mean square (rms) noise before and after integrating to 13 minute LST bins (top and bottom sets of curves) indicating the rising significance of foregrounds as we take the last deep integration step. Noise is estimated by differencing adjacent frequencies (magenta) and between redundant baselines (blue), compared with an estimate of the theoretical noise level with a $T_{sys}$ of 550K (dotted). The top three lines show noise after filtering foregrounds and binning into 40s long sidereal bins.  At this noise level the frequency and baseline differences are roughly similar, both demonstrate the same small bumps of increased noise due to interference flagging and are consistent with noise over much of the band.  The bottom three lines show the noise level after integrating up to the maximum fringe rate of 776s. The step between the top and bottom sets is the last coherent integration with noise largely decreasing by the expected factor $\sqrt{776/40}$ except in the difference between baselines (blue) which demonstrates a clear excess at all frequencies, particularly above redshift 10. With this last large jump in sensitivity we are now seeing the slight dominance of baseline covariance over the spectral derivative rms. }
\end{figure}

@@

\section{Results}
\label{sec:results}
\subsection{Foreground Filtering and Noise Levels}
\label{sec:noise}
The root-mean-square of the filtered visibilities measures how well we've removed foregrounds.  In Figure \ref{fig:noise} we examine $T_{rms}$ as calculated by differencing between adjacent channels and between redundant baselines. We see that in the last stage of coherent integration, the step between 40 second to 789 second integrations, the noise level decreases by the expected factor of $\sqrt{789/40}$.  At both stages the noise level found by both methods is roughly consistent with the theoretical level, only deviating significantly at the edges where the effective signal to noise due to the decreasing passband gain, Blackman Harris window, and uneven sampling (see Fig \ref{fig:bins}) drops precipitously.

The theoretical visibility r.m.s. noise level is a straightforward calculation, but given the method employed here, should be given due description.
The noise level in a long integration (Eq. \ref{eq:Trms}) is caused by the temperature of the sky (mostly due to galactic emission) which varies with LST and by the receiver noise, both on the scale of a few 100K. The noise of any long integration is also dependent on the number of observations in each LST.  The number of points in each bin is mostly determined by the observing schedule, which for PAPER is sunrise to sunset, but also by any flagging of data or observing outage (only 92 of the 103 days in the observing period have recorded data).  Unfortunately, technical problems limited the retention of the exact cumulative count of points averaged into each LST bin during the averaging process.  Here we have chosen to use the first order estimate based on the observing schedule to estimate the number of observations in each LST. Some bins are observed only a few times, while a slim range in the LST range 6 - 9 hours is observed on every night.  Over a number of LST bins $N_\textrm{lst}$ the average effective time per bin $N_{eff}$ is given by

%\[
%\frac{1}{\sqrt{N_{eff}}} = \frac{1}{N_{\textrm{lsts}}} \sum_\textrm{lst=1}^\textrm{lst=9}{\frac{1}{\sqrt{N(lst)}}}
%\]
\begin{equation}
\frac{1}{\sqrt{N_{eff}}}  = \left\langle \frac{1}{\sqrt{N(lst)}}  \right\rangle_{\textrm{lst}}
\end{equation}
On average each bin is in this dataset is effectively measured 39.6 times. Given this accounting of the number of samples $N(lst)$, we estimate the r.m.s. temperature of the averaged data set to be

%\[
%T_{rms}  =\frac{1}{N_{\textrm{lsts}}}\sum_{\textrm{lst}} \frac{T_{sky}(lst) + T_{rcvr}}{\sqrt{2BtN(lst)}} \approx \frac{1}{\sqrt{2BtN_{eff}}} \left(\frac{1}{N_{\textrm{lsts}}}\sum_{\textrm{lst}}T_{sky}(lst) + T_{rcvr}\right)
%\label{eq:Trms}
%\]

\begin{equation}
T_{rms}  =\frac{1}{N_{\textrm{lsts}}}\sum_{\textrm{lst}} \frac{T_{sky}(lst) + T_{rcvr}}{\sqrt{2BtN(lst)}} \approx \frac{(\langle T_{sky}\rangle + T_{rcvr})}{\sqrt{2BtN_{eff}}}
\label{eq:Trms}
\end{equation}

\noindent{where B and t are the 496kHz channel width and lst bin length, respectively and the approximation indicates that we have assumed average values for the number of samples and the time dependent sky component of the system temperature.  In P14 the total system temperature ($T_{sky} + T_{rcvr}$) was estimated to be 550K using 20 nights of the 92 night set which covered LSTs 3-9. Using the global sky model  \citep{deOliveiraCosta:2008p2242} and a model of the PAPER primary beam, we estimate the mean $T_{sky}$ during that 20 night period to be 250K, implying a receiver temperature of 300K.  The mean temperature over the full 92 night period is only 4K higher; the estimate of 550K for the system temperature is a good estimate for the full observing period.  $T_{rms}$ using 550K and an effective sample count of 39.6  is plotted as a dotted line in Figure \ref{fig:noise} for before and after fringe rate filtering (t=40s and 789s)}

The final fringe-rate step is the last coherent average of the processing routine and represents the last large gain in sensitivity; all following averaging steps are done incoherently on the square of the visibilities where sensitivity is gained at a much slower rate. For this reason the noise curves of the fringe-rate averaged data set (bottom set of curves of Figure \ref{fig:noise})  merit closer inspection.  At this sensitivity level we see a notable excess of power in the r.m.s. difference between baselines. Though the cause of this excess is unknown, it is suggestive of the excess covariance between redundant baselines which we remove in section \ref{sec:power_spectrum}, a hypothesis also supported by the high levels of residual foregrounds in the redshift 10.5 bin where the baseline difference r.m.s. is highest.

@@

If our measurements were completely noise dominated, $68\%$ of the data points ought to include zero within their error bars.  In Table \ref{tab:data}, we divide the points into ``Detections" (power spectrum values smaller than $1\sigma$ errors), and ``Upper limits" (measured values larger).  With only one exception, this scheme yields the same results whether we use the theoretical errors or the bootstrap errors.  Only a minority, $\sim$20\%, of points are consistent with noise at the 1 $\sigma$ level, and considering all the data points together, the k modes above 0.2 in Table \ref{tab:data} are inconsistent with zero at 6.6 $\sigma$.  This quantifies the positive bias that is visually apparent in Figures \ref{fig:pspecs} and \ref{fig:slices}.

%In general the theoretical error bars are similar to the bootstrap error bars.  This is probably most clear in Figure \ref{fig:slices} which is plotted on a linear scale and along the redshift axis where correlations between points should be minimal where, though most points are slightly outside the error region, the bootstrap errors do extend inside the noise dominated region. In general it would appear that the sensitivity calculations under-estimate the sensitivity of the measurements by a small factor. See \cite{Pober:2014p10390,Pober:2013p9581} for a discussion of the approximations made in those calculations.

Most of the power spectrum points have an excess power at about the 2$\sigma$ level.  As these measurements have a sensitivity level which is still two orders of magnitude away from theoretical reionization levels we consider the likelihood that these detections are evidence of reionization to be small. Far more likely is that we are limited by a residual foreground or systematic arising from approximations in the data analysis.  The statistical significance of the residual --2$\sigma$ after averaging over the entire season-- is low enough that it is not possible to break it down along likely axes like sidereal time or time of day to probe its origin.  Such studies must await higher sensitivity measurements with more antennas.  However, there are several additional features that seem to suggest an excess originating in foregrounds rather than a HI signal

%  That figure also provides an additional clue in the

%Generally, both figures suggest a power spectrum containing instrumental noise and foregrounds entering at comparable levels, with the latter possibly leaking from within the horizon to modes of higher $k$.  To see this and other systematic influences that may be present, one can compare the bootstrap error bars, the theoretical error bars, and the power spectrum values.    After a length of observation similar to what will be necessary to complete a detection level integration with the completed 128 element array, there are clear systematics, though at a low enough level that only a limited analysis of their origins is possible.

%Generally, both figures convey a picture of a power spectrum dominated by noise, with foregrounds encroaching from inside and just outside the horizon to higher $k$ ``reionization'' modes.  One easy way to see this leakage is by comparing the bootstrap error bar, the theoretical error and the power spectrum value.  At the level of noise here, we expect all points outside the horizon to be consistent with zero. As a simple sorting criteria we divide the points into ``Detections'' and ``Upper Limits'' (Det/ULim) in Table \ref{tab:data} for both theoretical and bootstrap error bars, finding agreement for all but one data point.  A minority $\sim$20\% of points are consistent with noise distributed about zero at 1 $\sigma$.  Taken together, all the points are inconsistent with zero at 6.6 $\sigma$.   %As we can see, the larger theoretical error bars are mostly consistent with zero, while the same is true of a little less then half of the bootstrap error bars.  This is consistent with a picture of systematics or foreground leakage near the thermal noise limit.

The predicted signature of the HI signal during reionization is one of a relatively short burst of power at small $k$ values as the ionized bubbles are briefly at their largest, before the signal from the ever increasing ionized medium drops sharply \citep{Pritchard:2008p8123}. In contrast, this observed residual is relatively constant with redshift while rising at both ends of the band, mimicking the RMS noise curve (Figure \ref{fig:noise}).  This suggests that the accuracy of the foreground filter decreases towards the edges of the band.  Recall that the wide-band foreground model we have subtracted was built by weighting the entire band by a Blackman-Harris window function which provides a much higher isolation of foregrounds inside the horizon. The tradeoff that is made is that data in the outer parts of the spectrum are heavily down weighted.  For example data at 126MHz are down-weighted $\sim$75\% compared to 150Mhz.  This amounts to making an assumption about the spectral smoothness of the data, namely that 126MHz can be mostly modeled by extrapolation from 150MHz.  The precision of this subtraction can be judged by examining the foreground residuals inside the horizon in the $P(k)$ power spectra plotted in the upper part of Figure \ref{fig:pspecs}\footnote{Note that this notch is not present in Figure \ref{fig:P14compare} because the foregrounds have been left in for comparison with P14.}.  The relative depth of the redshift 10.5 P(k) foreground residual is substantially shallower then at 7.55 and is generally correlated with the amount of residual found at higher $k$s.

The residual level is also illustrated in the $k$ slices of Figure \ref{fig:slices}.  
 The shortest $k$ mode (upper left of Figure \ref{fig:slices}) is the nearest to the light travel horizon and therefore most likely to be contaminated by foreground leakage. This $k$ mode is well above the noise and has a redshift dependence much the same shape as the $T_{rms}$ curve in Figure \ref{fig:noise}, with a minimum near redshift 8.5 and a dramatic rise above redshift 10. Longer $k$ modes are all generally positively biased, though none higher than 3 $\sigma$.

@@

\begin{deluxetable*}{llrrrrr}
\tablecolumns{7}
\tablecaption{Power spectrum values. Most are inconsistent with noise at $\sim$2$\sigma$, indicating that this full season of integration sees systematics at about twice the level of the noise. In sum the points above $k=0.2$ are inconsistent with zero at 6.6 $\sigma$.  On average the theoretical and bootstrap estimates of the noise agree to within 15\%.}
\tablehead{
\colhead{k} & \colhead{redshift}  & \colhead{$\Delta^2$} &
\colhead{bootstrap error } &\colhead{\parbox{6em}{noise model error} }& \colhead{\parbox{5em}{bootstrap significance}} & \colhead{\parbox{6em}{noise model significance}}\\
\colhead{[hMpc$^{-1}$]} & \colhead{}  & \colhead{[mK$^2$]} &
\colhead{ [mK$^2$]} &\colhead{\parbox{7em}{[mK$^2$]} }& \colhead{} & \colhead{}
}
\startdata
0.1 & 10.29 & 43575.6 & $\pm$ 5729.3 & $\pm$ 676.4 & Det & Det \tabularnewline
0.1 & 8.54 & 5701.6 & $\pm$ 1168.7 & $\pm$ 279.7 & Det & Det \tabularnewline
0.1 & 7.94 & 7695.4 & $\pm$ 2044.6 & $\pm$ 229.3 & Det & Det \tabularnewline
0.1 & 7.55 & 9461.8 & $\pm$ 2761.6 & $\pm$ 153.3 & Det & Det \tabularnewline
0.2 & 10.29 & 11411.0 & $\pm$ 5039.8 & $\pm$ 5326.1 & Det & Det \tabularnewline
0.2 & 8.54 & 5855.9 & $\pm$ 2195.8 & $\pm$ 2184.4 & Det & Det \tabularnewline
0.2 & 7.94 & 5143.7 & $\pm$ 1799.7 & $\pm$ 1785.6 & Det & Det \tabularnewline
0.2 & 7.55 & 856.3 & $\pm$ 1490.2 & $\pm$ 1190.0 & ULim & ULim \tabularnewline
0.3 & 10.29 & 34756.5 & $\pm$ 16384.8 & $\pm$ 17922.7 & Det & Det \tabularnewline
0.3 & 8.54 & 12923.3 & $\pm$ 5373.5 & $\pm$ 7339.6 & Det & Det \tabularnewline
0.3 & 7.94 & 5204.4 & $\pm$ 4834.5 & $\pm$ 5996.1 & Det & ULim \tabularnewline
0.3 & 7.55 & 7628.5 & $\pm$ 4810.3 & $\pm$ 3993.6 & Det & Det \tabularnewline
0.4 & 10.29 & 42910.0 & $\pm$ 30695.0 & $\pm$ 42439.6 & Det & Det \tabularnewline
0.4 & 8.54 & 24115.2 & $\pm$ 16585.6 & $\pm$ 17370.2 & Det & Det \tabularnewline
0.4 & 7.94 & 14730.7 & $\pm$ 9930.2 & $\pm$ 14187.9 & Det & Det \tabularnewline
0.4 & 7.55 & -8601.8 & $\pm$ 14191.5 & $\pm$ 9447.4 & ULim & ULim \tabularnewline
\enddata
\tablenotetext{a}{All error bars are 1 $\sigma$.}
\tablenotetext{b}{Det indicates a measurement and error-bar inconsistent with zero, ULim indicates consistency with zero at 1 $\sigma$.}
\label{tab:data}
\end{deluxetable*}

>>>>

\documentclass[fleqn,10pt,iop]{emulateapj}
%\documentclass[preprint2]{aastex}

@@

\tabletypesize{\scriptsize}

@@

\title{Multi-redshift limits on the 21cm power spectrum from PAPER}
\author{
Daniel C. Jacobs\altaffilmark{1},
Jonathan C. Pober\altaffilmark{12},
Aaron R. Parsons\altaffilmark{2,8},
James E. Aguirre\altaffilmark{3},
Zaki Ali\altaffilmark{2},
Judd Bowman\altaffilmark{1},
Richard F. Bradley\altaffilmark{4,5,6},
Chris L.  Carilli\altaffilmark{7,10},
David R. DeBoer\altaffilmark{8},
Matthew R. Dexter\altaffilmark{8},
Nicole E. Gugliucci\altaffilmark{5},
Pat Klima\altaffilmark{5},
Adrian Liu\altaffilmark{2,11},
Dave H. E. MacMahon\altaffilmark{8}
Jason R. Manley\altaffilmark{9},
David F. Moore\altaffilmark{3},
Irina I. Stefan\altaffilmark{10},
William P. Walbrugh\altaffilmark{9}}

@@

\begin{abstract}
The epoch of reionization power spectrum is expected to evolve strongly with redshift, and it is this variation with cosmic history that will allow us to begin to place constraints on the physics of reionization.  The primary obstacle to the measurement of the EoR power spectrum is bright foreground emission.  We present an analysis of observations from the Donald C. Backer Precision Array for Probing the Epoch of Reionization (PAPER) telescope which place new limits on the HI power spectrum over the redshift range of $7.5<z<10.5$, extending previously published single redshift results to cover the full range accessible to the instrument.   To suppress foregrounds, we use filtering techniques that take advantage of the large instrumental bandwidth to isolate and suppress foreground leakage into the interesting regions of $k$-space.  Our 500 hour integration is the longest such yet recorded and demonstrates this method to a dynamic range of $10^4$.   Power spectra at different points across the redshift range reveal the variable efficacy of the foreground isolation.  Noise limited measurements of $\Delta^2$ at $k=$0.2hMpc$^{-1}$ and z$=7.55$ reach as low as (48mK)$^2$ ($1\sigma$).  We demonstrate that the size of the error bars in our power spectrum measurement as generated by a bootstrap method is consistent with the fluctuations due to thermal noise.  Relative to this thermal noise, most spectra exhibit and excess of power at a few sigma.  The likely sources of this excess include residual foreground leakage, particularly at the highest redshift, and unflagged RFI.   We conclude by discussing data reduction improvements that promise to remove much of this excess. 
\end{abstract}

@@

 \begin{figure*}
\centering
\includegraphics[width=\textwidth]{figures/psa32_flagging_zbins.png}
\caption{\label{fig:bins} The average amount of data remaining after interference flagging over the 3 month period between Dec 2011 and March 2012 (black line) is quite high.  Redshift bins (in grey, redshift center indicated with text label) are chosen to include spectral channels with uniform weight, i.e. no missing channels while maximizing coverage over the band.  Redshift 7.68 is included for comparison with P14.  Channels with no missing data are indicated by the dotted line, visible at the edge of flagged channels. Each redshift bin is 20 MHz wide, but weighted by a Blackman-Harris window function which heavily down-weights the outer 10MHz for a Noise Equivalent Bandwidth of 10MHz.  The interference is almost exclusively dominated by two features: ORBCOMM satellites at 137MHz and an unidentified intermittent line emitter at 175MHz. The roll off at 115MHz is due the rising noise at low frequencies being incorrectly flagged as interference.  }
\end{figure*} 
%The success of the remainder of the processing stages hinges on the spectral stability of the data, which can easily be assessed by comparing redundant baselines. In figure XXX we see the r.m.s. of the ratios between redundant calibrations, before and after the calibration has been applied.  
\begin{figure}
\centering
\includegraphics[width=\columnwidth]{figures/psa32_trms_with_and_without_fr_filtering_z.png}
\caption{\label{fig:noise} Root mean square (rms) noise before and after integrating to 13 minute LST bins (top and bottom sets of curves) indicating the rising significance of foregrounds as we take the last deep integration step. Noise is estimated by differencing adjacent frequencies (magenta) and between redundant baselines (blue), compared with an estimate of the theoretical noise level with a $T_{sys}$ of 550K (dotted). The top three lines show noise after filtering foregrounds and binning into 40s long sidereal bins.  At this noise level the frequency and baseline differences are roughly similar, both demonstrate the same small bumps of increased noise due to interference flagging and are consistent with noise over much of the band.  The bottom three lines show the noise level after integrating up to the maximum fringe rate of 789s. The step between the top and bottom sets is the last coherent integration with noise largely decreasing by the expected factor $\sqrt{789/40}$ except in the difference between baselines (blue) which demonstrates a clear excess at all frequencies, particularly above redshift 10. With this last large jump in sensitivity we are now seeing the slight dominance of baseline covariance over the spectral derivative rms. }
\end{figure}

@@

\section{Results}
\label{sec:results}
\subsection{Foreground Filtering and Noise Levels}
\label{sec:noise}
The root-mean-square of differenced visibilities measures how well we've removed foregrounds.  In Figure \ref{fig:noise} we examine $T_{rms}$ as calculated by differencing between adjacent channels and between redundant baselines.  The square of this difference is then averaged over the full LST range probed by the power spectra (1h-9h). Note that this operation combines all LST modes as though they were repeated measurements of the same sky when in practice the power spectrum values are averaged in quadrature over different LST bins. This provides a lower bound on the sensitivity possible in the data set and a sensitive probe on foreground residuals or other sources of excess power.  Using all the data in this way, we expect to illuminate rare outliers that effect the bottom line.

\subsubsection{The Theoretical RMS Noise Level}

The expected noise level over this time range depends on the number of samples in each lst bin and the system temperature in that bin. System noise comes in part from the bright radio sky and part from amplifier noise in the receiver. As well as the channel width $B$ and the integration time $t$

\begin{equation}
T_{rms}  =\frac{1}{N_{\textrm{lsts}}}\sum_{\textrm{lst}} \frac{T_{sky}(lst) + T_{rcvr}}{\sqrt{2BtN(lst)}} 
\label{eq:Trms}
\end{equation}

Where $T_{sky}(lst)$ is modeled by the convolution of the global sky model \citep{deOliveiraCosta:2008p2242} with the PAPER beam \citep{Pober:2012p8800} which for the LST range here averages to 250 $(\nu/160)^(-2.6)$~K.

One of the strongest influences on the sensitivity of wide-field, low frequency instruments like PAPER or the MWA is the number of samples local sidereal time bin available in an observing year.   After removing data containing Sun and the Galactic center, both of which are sources of strong foreground emission, the amount of time required to record a deep $\sim$1000 hour observation can stretch into multiple years.  Some LST bins in this range will never be sampled more than a few times per year.  This observation, which ran from December to March, measured samples over half the night sky. However only the range 6h42 to 9h was measured every night to reach ``full'' sensitivity of the three-month-long observing season.  Integrating equation \ref{eq:Trms} over this sampling function we find a $T_{rms}$ noise level equivalent to 40 nights of integration.

Using the variance of a 20 day subset, P14 found the system temperature for this LST range to be 550~K at 160~MHz, while the sky model suggests a $T_{sky}$ of 250~K.  This level of sky temperature is confirmed by the fit of the theoretical noise level in Figure \ref{fig:noise}.   The sky portion of the noise has a significant slope. Ascribing more then 250~K of the total noise to the sky results in a significantly steeper theoretical $T_{rms}$ in Figure \ref{fig:noise}. Taken together this evidence points to a receiver noise temperature of 300~K.

In summary we find that the noise level of this data set is well modeled by a 550~K system temperature which is nearly equal parts sky and receiver and when integrated over our lst sampling function is equivalent to 40 days of data.

\subsubsection{Analysis of the RMS Noise Level}

Here we have estimated the residual noise after foreground filtering and LST averaging, both before and after the last ``coherent'' integration step, when fringe rate filtering increases the effective integration time from 40 to 789 seconds.  
The final fringe-rate step is the last coherent average of the processing routine and represents the last large gain in sensitivity; all following averaging steps are done incoherently on the square of the visibilities where sensitivity is gained at a much slower rate.  The noise level of the 40s average (top curves of Figure \ref{fig:noise}) is consistent between the two methods of noise estimation and with the predicted level.

After the last stage of coherent integration (from the top set of curves to the bottom set), where the integration increases from 40 to 789 second integrations.  At this deepest level of integration the noise levels rise slightly above the expected decrease of $\sqrt{789/40}$. In both cases the noise curves deviate significantly from the theory at the edges of the band. This is mainly due to our neglect of variable sensitivity across the band (e.g. see Figure \ref{fig:bins}) in calculating the theoretical curve; note that the band edges still integrate down between the two steps. However it does not integrate down perfectly, residual levels are apparent across the entire band with larger excesses at the band edges.  The largest excess is noted in the rms difference between redundant baselines.  Lack of agreement between supposedly redundant baselines is suggestive of a breakdown in our assumption of spectrally and temporally smooth calibration. It is also suggestive of the excess covariance between redundant baselines which we remove in section \ref{sec:power_spectrum}, a hypothesis also supported by the high levels of residual foregrounds in the redshift 10.5 bin where the baseline difference r.m.s. is highest.

@@

If our measurements were completely noise dominated, $68\%$ of the data points ought to include zero within their error bars.  In Table \ref{tab:data}, we divide the points into ``Detections" (power spectrum values smaller than $1\sigma$ errors), and ``Upper limits" (measured values larger).  With only one exception, this scheme yields the same results whether we use the theoretical errors or the bootstrap errors.  Under this scheme, we would expect a noise dominated signal to yield around eight ``upper limits'' with an equal number of positive and negative data points. Here we find only two to three upper limits while rest we classify as detections. In general, most detections are within 1 - 2 $\sigma$ but are all positively biased,  with the significance rising steeply at high redshifts and low $k$s.%, the quadratic sum of all errors above $k=0.1$ is 6.6 $\sigma$ indicating that the excess is largely coherent across the data set.

\subsection{The Excess Signal}

As these measurements have a maximum possible sensitivity level which is still two orders of magnitude away from theoretical reionization levels we consider the likelihood that these detections are evidence of reionization to be small. Far more likely is that we are limited by a residual foreground or systematic arising from approximations in the data analysis.  The statistical significance of the residual --2$\sigma$ after averaging over the entire season-- is low enough that it is not possible to break it down along likely axes like sidereal time or time of day to probe its origin.  Such studies must await higher sensitivity measurements with more antennas.  However, there are several additional features that seem to suggest an excess not caused by high redshift HI.

The spectral signature of the measured excess is far more similar to a foreground residual then to theoretical HI models.
The predicted signature of the HI signal during reionization is one of a relatively short burst of power at small $k$ values as the ionized bubbles are briefly at their largest, before the signal from the ever increasing ionized medium drops sharply \citep{Pritchard:2008p8123}. In contrast, this observed residual is relatively constant with redshift while rising at both ends of the band, mimicking the RMS noise curve in Figure \ref{fig:noise}.  This suggests that the accuracy of the foreground filter decreases towards the edges of the band.  Recall that the wide-band foreground model we have subtracted was built by weighting the entire band by a Blackman-Harris window function which provides a much higher isolation of foregrounds inside the horizon. The tradeoff that is made is that data in the outer parts of the spectrum are heavily down weighted.  For example data at 126MHz are down-weighted $\sim$75\% compared to 150Mhz.  This amounts to making an assumption about the spectral smoothness of the data, namely that 126MHz can be mostly modeled by extrapolation from 150MHz.  The precision of this subtraction can be judged by examining residuals inside the horizon in the $P(k)$ power spectra plotted in the upper part of Figure \ref{fig:pspecs}\footnote{Note that this notch is not present in Figure \ref{fig:P14compare} because the foregrounds have been left in for comparison with P14.}.  The relative depth of the redshift 10.5 P(k) foreground residual is substantially shallower then at 7.55 and is generally correlated with the amount of residual found at higher $k$s.

The residual level is also illustrated in the $k$ slices of Figure \ref{fig:slices}.  
 The shortest $k$ mode (upper left of Figure \ref{fig:slices}) is the nearest to the light travel horizon and therefore most likely to be contaminated by foreground leakage. This $k$ mode is well above the noise and has a redshift dependence much the same shape as the $T_{rms}$ curve in Figure \ref{fig:noise}, with a minimum near redshift 8.5 and a dramatic rise above redshift 10. These points are also the only ones where error bars significantly deviate from their predicted thermal values. As we see in Table \ref{tab:data}, going by predicted noise alone points closest to the horizon have error bars in excess of 20 $\sigma$ while the bootstrap error bars register a much larger uncertainty.  This hints at a more complicated origin for the excess. If these points were purely foregrounds, the error bars would be noise dominated, the fact that the excess is not much larger then its variation suggests a source of error not accounted in our analysis. Likely culprits include incomplete flagging of interference, instrumental glitches, or time of frequency dependent calibration variation which introduces differences between redundant baselines. All of which have are part of ongoing analysis efforts which are briefly discussed in the conclusions.   All of these effects have the effect of adding time and frequency dependence to the bright foregrounds limiting the effectiveness of the filter.   Thus, just like foregrounds, the strong excess drops rapidly towards higher $k$ modes. Above $k$ of 0.1 hMpc$^{-1}$  the power spectra are generally positively biased, though most close to 2 $\sigma$ or less using either theoretical and bootstrap error bars.

@@

\begin{deluxetable}{llccccccc}
\tablecolumns{9}
\tablecaption{PAPER Power spectrum values using 32 antenna and a full season of integration.}
\tablehead{
\colhead{k} & 
\colhead{redshift}  & 
\colhead{$\Delta^2$} &
\colhead{\parbox{6em}{theoretical error} }&
\colhead{\parbox{6em}{theoretical significance}} &
\colhead{\parbox{6em}{theoretical detection}} &
\colhead{\parbox{6em}{bootstrap error }} &
\colhead{\parbox{6em}{bootstrap significance}} &
 \colhead{\parbox{5em}{bootstrap detection}} \\
\colhead{[hMpc$^{-1}$]} & 
\colhead{}  & 
\colhead{[mK$^2$]} &
\colhead{ $\sigma$} & 
\colhead{}&
\colhead{\parbox{7em}{[mK$^2$]} }& 
\colhead{[mK$^2$]}&
\colhead{$\sigma$} & 
\colhead{}
}
\startdata
0.1 & 10.29 & 43600 & $\pm$   700 & 64.4 & Det & $\pm$  5700 & 7.6 & Det \tabularnewline
0.1 & 8.54 &  5700 & $\pm$   300 & 20.4 & Det & $\pm$  1200 & 4.9 & Det \tabularnewline
0.1 & 7.94 &  7700 & $\pm$   200 & 33.6 & Det & $\pm$  2000 & 3.8 & Det \tabularnewline
0.1 & 7.55 &  9500 & $\pm$   200 & 61.7 & Det & $\pm$  2800 & 3.4 & Det \tabularnewline
0.2 & 10.29 & 11400 & $\pm$  5300 & 2.1 & Det & $\pm$  5000 & 2.3 & Det \tabularnewline
0.2 & 8.54 &  5900 & $\pm$  2200 & 2.7 & Det & $\pm$  2200 & 2.7 & Det \tabularnewline
0.2 & 7.94 &  5100 & $\pm$  1800 & 2.9 & Det & $\pm$  1800 & 2.9 & Det \tabularnewline
0.2 & 7.55 &   900 & $\pm$  1200 & 0.7 & ULim & $\pm$  1500 & 0.6 & ULim \tabularnewline
0.3 & 10.29 & 34800 & $\pm$ 17900 & 1.9 & Det & $\pm$ 16400 & 2.1 & Det \tabularnewline
0.3 & 8.54 & 12900 & $\pm$  7300 & 1.8 & Det & $\pm$  5400 & 2.4 & Det \tabularnewline
0.3 & 7.94 &  5200 & $\pm$  6000 & 0.9 & ULim & $\pm$  4800 & 1.1 & Det \tabularnewline
0.3 & 7.55 &  7600 & $\pm$  4000 & 1.9 & Det & $\pm$  4800 & 1.6 & Det \tabularnewline
0.4 & 10.29 & 42900 & $\pm$ 42400 & 1.0 & Det & $\pm$ 30700 & 1.4 & Det \tabularnewline
0.4 & 8.54 & 24100 & $\pm$ 17400 & 1.4 & Det & $\pm$ 16600 & 1.5 & Det \tabularnewline
0.4 & 7.94 & 14700 & $\pm$ 14200 & 1.0 & Det & $\pm$  9900 & 1.5 & Det \tabularnewline
0.4 & 7.55 & -8600 & $\pm$  9400 & -0.9 & ULim & $\pm$ 14200 & -0.6 & ULim \tabularnewline
\enddata
\tablenotetext{a}{All error bars are 1 $\sigma$.}
\tablenotetext{b}{Det indicates a measurement and error-bar inconsistent with zero, ULim indicates consistency with zero at 1 $\sigma$.}
\label{tab:data}
\end{deluxetable}
